---
title: "Cleaning & Wrangling"
format: html
editor: visual
---

```{r}
#| message: false
#| echo: false

library(dplyr)
library(tidyr)
library(readr)
```

## Read in our data
```{r}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore the data
```{r}
colnames(catch_original)

head(catch_original, 3)

summary(catch_original)

unique(catch_original$Region)
```

# Clean and Reshape data
# Ctrl + Shift + M for pipe operator "%>%"
# Pipe operator instructs the next line to use that line's output in its run

## Using select() function
```{r}
catch_data <- catch_original %>%
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum) # Choose variables
head(catch_data)

catch_data <- catch_original %>%
  select(-All, -notesRegCode) # Drop variables
head(catch_data)
```

# Check column types
```{r}
glimpse(catch_data)
```


## Replace values in a column
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = as.numeric(Chinook))
```

## Find NA values
```{r}
i <- which(is.na(catch_clean$Chinook))
i

catch_data[i, ]
```

## Change NA value to integer
```{r}
catch_clean <- catch_data %>%
    mutate(Chinook = if_else(condition = Chinook == "I", 
                             true = "1", 
                             false = Chinook),
           Chinook = as.numeric(Chinook))

## Check
catch_clean[i, ]
# Ensure that catch_clean is also fixed
```

## Reshape in long format
```{r}
catch_long <- catch_clean %>% 
    #pivot longer all columns except Region and Year
    pivot_longer(
        cols = -c(Region, Year),
        names_to = "species",
        values_to = "catch"
    )

head(catch_clean)
head(catch_long)
nrow(catch_long)
```

## Reshape in wide format
```{r}
catch_wide <- catch_long %>%
    pivot_wider(names_from = species,
                values_from = catch)

head(catch_long, 3)
head(catch_wide, 3)
nrow(catch_wide)
```

# Rename misleading column name
```{r}
catch_long <- catch_long %>%
    rename(catch_thousands = catch)

head(catch_long)
```

# Add column using mutate() and remove catch_thousands
```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)
head(catch_long)

catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000) %>%
    select(-catch_thousands)
head(catch_long)
```

# Summary statistics using group_by() and summarize()
```{r}
unique(catch_original$Region)

# Get average catch by region
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)

# Get number of rows in each group
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```

# Filtering rows using filter() function
```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)
```

# Filter to catches >one million
```{r}
millionPlus_catch <- catch_long %>%
    filter(catch > 1000000)

head(millionPlus_catch)
```

# Filter to just Chinook from the SSE region
```{r}
sse_chinook <- catch_long %>%
    filter(Region == "SSE" & species == "Chinook")

head(sse_chinook)
```

# Sort data using arrange()
```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch))

head(mean_region)
```

# Splitting a column using separate() and unite()
```{r}
sites_df <- data.frame(site = c("HAWLEY-101",
                                "HAWLEY-103",
                                "OAHOUI-320",
                                "OAHOUI-219",
                                "MAUNUA-039"))

sites_df %>%
    separate(site, c("island", "site_number"), "-")
```

# Split the city column in the data frame cities_df into city and state_code columns
```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))

cities_clean <- cities_df %>%
    separate(city, c("city", "state_code"), " ")

cities_clean
```

# Combine dataframes
```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)

dates_df %>%
    unite(date, year, month, day, sep = "-")
```

# Combine everything
```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

