---
title: "Introduction to Quarto"
format: html
bibliography: references.bib
---

# Cool Data Visualization

A simple demonstration of using Quarto to mix narrative and R code for data visualization.

## Distribution of the **`mtcars`** Dataset

Let's *visualize* the distribution of miles per gallon (mpg) in the `mtcars` dataset.

-   Shopping list
-   Item 1
-   Item 2

```{r}
# Load necessary libraries
library(ggplot2)
library(plotly)

# Create a histogram with ggplot2
p <- ggplot(mtcars, aes(x = mpg)) + 
     geom_histogram(binwidth = 1, fill = "lightgreen", color = "purple") +
     theme_minimal() +
     ggtitle("Distribution of Miles Per Gallon in mtcars Dataset")

# Convert the ggplot object to a plotly object
ggplotly(p)
```

```{r}
x <- 4 * 8
x
hights_ft <- c(5.2, 6.0, 5.7)

coef <- 3.14
```

```{r}
# Define the parts of the string
part1 <- "Ar"
part2 <- "ct"
part3 <- "ic"

# Concatenate the parts to form the string "Arctic"
arctic_string <- paste0(part1, part2, part3)

# Print the result
arctic_string
```



**Help -\> Cheat Sheets -\> R Markdown Cheat Sheet/Reference Guide**

<a href="https://cryoilrj.me" target="_blank">Visit my website ❄️</a>

Here is a sample paper: [@kufner2021]

| Penguin | Walrus | Orca |
|---------|--------|------|
| 1       | 2      | 4    |
| 4       | 5      | 2    |
| 20      | 1      | 7    |
